---
title: "Two-Stage YOLO Object Detection and Classification Model Report for Diptera Wings"
author: "Aaron Harmer"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

# Load required libraries
library(knitr)
library(ggplot2)
library(readr)
library(dplyr)
library(tidyr)
library(gridExtra)
library(corrplot)
library(RColorBrewer)
library(plotly)
library(DT)
library(scales)

# Set theme for consistent plotting
theme_set(theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    axis.title = element_text(size = 11),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10)
  ))

# Define paths - UPDATE THESE TO MATCH YOUR STRUCTURE
# Detection model paths
detection_model_dir <- "D:/Dropbox/data/dipteraID/runs/detect/model11n_ep30_autobatch"
detection_results_dir <- file.path(detection_model_dir, "results")

# Classification model paths  
classification_model_dir <- "D:/Dropbox/data/dipteraID/runs/classify/model-11n-cls_ep30_autobatch_do02_lr001"
classification_results_dir <- file.path(classification_model_dir, "results")

# Dataset paths
detection_dataset_path <- "D:/Dropbox/data/dipteraID/imgs/detection_set"
classification_dataset_path <- "D:/Dropbox/data/dipteraID/imgs/classification_set"
```

This report presents the results of a two-stage YOLO model system designed for dipteran identification. The two-stage approach allows for robust diptera wing detection followed by accurate generic classification.

## Methodology

### Model Architecture

The system employs a two-stage approach:

**Stage 1: Object Detection**
- Model: YOLOv11n (nano version for efficiency)
- Task: Detect diptera wing objects in images
- Output: Bounding boxes around detected wings
- Single-class detection (all wings treated as one class)

**Stage 2: Classification** 
- Model: YOLOv11n-cls (classification variant)
- Task: Classify detected diptera wing regions into genus
- Input: Cropped wing images from detection stage
- Output: generic classification with confidence scores

### Training Configuration

```{r training_config}
# Training configuration summary
training_config <- data.frame(
  Stage = c("Detection", "Classification"),
  Model = c("YOLOv11n", "YOLOv11n-cls"),
  Epochs = c(30, 30),
  Batch_Size = c(64, 92),
  Image_Size = c(640, 640),
  Optimizer = c("AdamW", "AdamW"),
  Dropout = c("N/A", 0.2),
  Single_Class = c("Yes", "No")
)

kable(training_config, 
      caption = "Training configuration for each model stage",
      align = "l")
```

All images were pre-processed before training by conversion to greyscale (there is no significant colour information in wing images), resizing to 640x640 pixels and adjusting the exposure by contrast stretching.

The object detection image set was augmented by flipping each wing image. The classification imageset was further augmented by generating random amounts of Gaussian noise and blur. The 'other' (hymenoptera) class was subsampled to be twice the size of the largest dipteran class, as it was much larger and likely to bias training.

## Dataset Analysis

### Detection Dataset Statistics

```{r detection_dataset_stats}
# Load detection dataset statistics
det_overall <- read_csv(file.path(detection_results_dir, "test_overall_metrics.csv"))
det_per_class <- read_csv(file.path(detection_results_dir, "test_per_class_metrics.csv"))

# Dataset size information
detection_stats <- data.frame(
  Split = c("Train", "Validation", "Test"),
  Images = c(det_per_class$Train_Images[1], 
             det_per_class$Val_Images[1], 
             det_per_class$Test_Images[1]),
  Ratio = c(label_percent(accuracy = 1)(det_per_class$Train_Images[1]/sum(det_per_class$Train_Images[1], det_per_class$Val_Images[1], det_per_class$Test_Images[1])), label_percent(accuracy = 1)(det_per_class$Val_Images[1]/sum(det_per_class$Train_Images[1], det_per_class$Val_Images[1], det_per_class$Test_Images[1])), label_percent(accuracy = 1)(det_per_class$Test_Images[1]/sum(det_per_class$Train_Images[1], det_per_class$Val_Images[1], det_per_class$Test_Images[1])))
)

kable(detection_stats, 
      caption = "Detection dataset distribution",
      align = "l")

```

### Classification Dataset Statistics

```{r classification_dataset_stats, results='asis'}
# Load classification dataset statistics
cls_overall <- read_csv(file.path(classification_results_dir, "classification_overall_metrics_test.csv"))
cls_class_metrics <- read_csv(file.path(classification_results_dir, "classification_class_metrics_test.csv"))

# Number of classes
num_classes <- nrow(cls_class_metrics)

cat("**Number of generic classes:**", num_classes, "\n\n")

# Summary statistics
total_stats <- data.frame(
  Split = c("Train", "Validation", "Test"),
  Images = c(
    sum(cls_class_metrics$`Train Images`, na.rm = TRUE),
    sum(cls_class_metrics$`Val Images`, na.rm = TRUE),
    sum(cls_class_metrics$`Test Images`, na.rm = TRUE)),
  Ratio = c(label_percent(accuracy = 1)(cls_class_metrics$"Train Images"[1]/sum(cls_class_metrics$"Train Images"[1], cls_class_metrics$"Val Images"[1], cls_class_metrics$"Test Images"[1])), label_percent(accuracy = 1)(cls_class_metrics$"Val Images"[1]/sum(cls_class_metrics$"Train Images"[1], cls_class_metrics$"Val Images"[1], cls_class_metrics$"Test Images"[1])), label_percent(accuracy = 1)(cls_class_metrics$"Test Images"[1]/sum(cls_class_metrics$"Train Images"[1], cls_class_metrics$"Val Images"[1], cls_class_metrics$"Test Images"[1])))
)

kable(total_stats,
      caption = "Classification dataset distribution",
      align = "l")

```

## Model Performance Results

### Detection Model Performance

```{r detection_performance}
  # Overall detection metrics
kable(det_overall, 
      caption = "Object detection performance metrics",
      digits = 2,
      align = "l")

# Per-class metrics (if multiple classes)
if(nrow(det_per_class) > 1) {
  det_display <- det_per_class %>%
    select(Class, `AP@0.50`, `AP@0.50:0.95`) %>%
    arrange(desc(`AP@0.50`))
  
  kable(det_display,
        caption = "Detection performance by class",
        digits = 2,
        align = "l")
}

```

### Classification Model Performance

```{r classification_performance, results='asis'}
# Overall classification metrics
overall_summary <- data.frame(
  Metric = c("Accuracy", "Precision", "Recall", "F1 Score"),
  Value = c(
    cls_overall$overall_accuracy,
    cls_overall$overall_precision, 
    cls_overall$overall_recall,
    cls_overall$overall_f1_score
  )
)

kable(overall_summary,
      caption = "Overall classification performance metrics",
      digits = 2,
      align = "l")

# Per-class classification metrics
cls_display <- cls_class_metrics %>%
  select(`Class Name`, `Test Images`, Accuracy, Precision, Recall, `F1 Score`) %>%
  arrange(desc(Accuracy))

kable(cls_display,
      caption = "Classification performance by genus",
      digits = 2,
      align = "l")

```

## Training Curves

### Detection Model Training Curves

```{r detection_training_curves, fig.height=8}
# Load training results - YOLO saves results.csv in the training directory
det_results_path <- file.path(detection_model_dir, "results.csv")

if(file.exists(det_results_path)) {
  det_training <- read_csv(det_results_path)
  
  # Clean column names (remove whitespace)
  names(det_training) <- trimws(names(det_training))
  
  measures = as_labeller(c("Loss" = "Training and Validation Loss", "metrics/mAP50-95(B)" = "mAP50-95"))
  
  det_training %>% 
    select("epoch", "train/box_loss", "val/box_loss", "metrics/mAP50-95(B)") %>% 
    pivot_longer(!epoch, names_to = "measure") %>% 
    # Create a grouping variable to combine the loss metrics
    mutate(panel = case_when(
      measure %in% c("train/box_loss", "val/box_loss") ~ "Loss",
      measure == "metrics/mAP50-95(B)" ~ "metrics/mAP50-95(B)"
    )) %>%
    # Create a line type/color variable for the loss panel
    mutate(line_type = case_when(
      measure == "train/box_loss" ~ "Training",
      measure == "val/box_loss" ~ "Validation",
      TRUE ~ NA_character_  # Set to NA for mAP50-95
    )) %>%
    ggplot(aes(x = epoch, y = value)) +
    # Apply color and linetype aesthetics only for non-NA line_type
    geom_line(aes(color = line_type, linetype = line_type), 
              data = . %>% filter(!is.na(line_type)), linewidth = 1) +
    geom_point(aes(color = line_type), 
               data = . %>% filter(!is.na(line_type)), size = 2) +
    # Add mAP50-95 line without aesthetics (so it won't appear in legend)
    geom_line(data = . %>% filter(is.na(line_type)), 
              color = "#2274b7", linewidth = 1) +
    geom_point(data = . %>% filter(is.na(line_type)), 
               color = "#2274b7", size = 2) +
    facet_wrap(~panel, labeller = measures, scales = "free_y") +
    scale_color_manual(values = c("Training" = "#2274b7", "Validation" = "lightblue"), 
                       na.translate = FALSE) +
    scale_linetype_manual(values = c("Training" = "solid", "Validation" = "solid"), 
                          na.translate = FALSE) +
    xlab("\nEpoch") +
    ylab("Value\n") +
    labs(color = "Metric", linetype = "Metric") +
    theme_minimal() +
    theme(panel.border = element_rect(fill = NA, linewidth = 1), 
          text = element_text(size = 14),
          legend.position = "bottom")
  
} else {
  cat("Detection training results.csv not found at:", det_results_path)
}

```

### Classification Model Training Curves

```{r classification_training_curves, fig.height=8}

# Load classification training results
cls_results_path <- file.path(classification_model_dir, "results.csv")

if(file.exists(cls_results_path)) {
  cls_training <- read_csv(cls_results_path)
  
  # Clean column names
  names(cls_training) <- trimws(names(cls_training))
  
  measures = as_labeller(c("Loss" = "Training and Validation Loss", "Top1 accuracy" = "Top1 accuracy"))
  
  cls_training %>% 
    select("epoch", "train/loss", "val/loss", "metrics/accuracy_top1") %>% 
    pivot_longer(!epoch, names_to = "measure") %>% 
    # Create a grouping variable to combine the loss metrics
    mutate(panel = case_when(
      measure %in% c("train/loss", "val/loss") ~ "Loss",
      measure == "metrics/accuracy_top1" ~ "Top1 accuracy"
    )) %>%
    # Create a line type/color variable for the loss panel
    mutate(line_type = case_when(
      measure == "train/loss" ~ "Training",
      measure == "val/loss" ~ "Validation",
      measure == "metrics/accuracy_top1" ~ "Top1 accuracy"
      # TRUE ~ NA_character_  # Set to NA for mAP50-95
    )) %>%
    ggplot(aes(x = epoch, y = value)) +
    # Apply color and linetype aesthetics only for non-NA line_type
    geom_line(aes(color = line_type, linetype = line_type), 
              data = . %>% filter(!is.na(line_type)), linewidth = 1) +
    geom_point(aes(color = line_type), 
               data = . %>% filter(!is.na(line_type)), size = 2) +
    # Add top1 line without aesthetics (so it won't appear in legend)
    geom_line(data = . %>% filter(line_type == "Top1 accuracy"), 
              color = "#2274b7", linewidth = 1) +
    geom_point(data = . %>% filter(line_type == "Top1 accuracy"), 
               color = "#2274b7", size = 2) +
    facet_wrap(~panel, labeller = measures, scales = "free_y") +
    scale_color_manual(values = c("Training" = "#2274b7", "Validation" = "lightblue", "Top1 accuracy" = "#2274b7")) +
    scale_linetype_manual(values = c("Training" = "solid", "Validation" = "solid", "Top1 accuracy" = "solid")) +
    xlab("\nEpoch") +
    ylab("Value\n") +
    labs(color = "Metric", linetype = "Metric") +
    theme_minimal() +
    theme(panel.border = element_rect(fill = NA, linewidth = 1), 
          text = element_text(size = 14),
          legend.position = "bottom")
  
  
} else {
  cat("Classification training results.csv not found at:", cls_results_path)
}

```

## Confusion Matrix and Classification Analysis

```{r confusion_matrix, fig.width=12, fig.height=8}

# Load detailed classification results
cls_detailed_path <- file.path(classification_results_dir, "classification_results_test.csv")

if(file.exists(cls_detailed_path)) {
  cls_detailed <- read_csv(cls_detailed_path)
  
  # Create confusion matrix
  confusion_data <- table(cls_detailed$true_label, cls_detailed$predicted_label)
  confusion_df <- as.data.frame(confusion_data)
  names(confusion_df) <- c("True_Label", "Predicted_Label", "Count")
  
  # Get class names from the class metrics
  # if(exists(cls_class_metrics)) {
    class_mapping <- cls_class_metrics %>%
      select("Class Index", "Class Name") %>%
      distinct()
    
    confusion_df <- confusion_df %>%
      left_join(class_mapping %>% mutate(`Class Index` = as.factor(`Class Index`)), 
                by = c("True_Label" = "Class Index")) %>%
      rename(True_Class = `Class Name`) %>%
      left_join(class_mapping %>% mutate(`Class Index` = as.factor(`Class Index`)), 
                by = c("Predicted_Label" = "Class Index")) %>%
      rename(Predicted_Class = `Class Name`) %>%
      select(True_Class, Predicted_Class, Count)
  # }
  
  
  con_mat = ggplot(confusion_df, aes(x = Predicted_Class, y = True_Class, fill = Count)) +
    geom_tile(color = "white") +
    geom_text(aes(label = ifelse(Count == 0, "", Count)), color = "black", size = 3) +
    scale_fill_gradient(low = "white", high = "steelblue") +
    labs(title = "Confusion Matrix - Genus Classification",
         x = "Predicted Genus", y = "True Genus") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  print(con_mat)
  
  
} else {
  cat("Detailed classification results not found at:", cls_detailed_path)
}

```

## Confidence Score Analysis

```{r confidence_analysis, fig.width=12}
if(exists("cls_detailed")) {
  # Confidence distribution
  p6 <- ggplot(cls_detailed, aes(x = confidence)) +
    geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
    labs(title = "Distribution of prediction confidence scores",
         x = "Confidence Score", y = "Frequency")
  
  # Confidence by correctness
  cls_detailed$correct <- cls_detailed$true_label == cls_detailed$predicted_label
  
  p7 <- ggplot(cls_detailed, aes(x = confidence, fill = correct)) +
    geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
    scale_fill_manual(values = c("FALSE" = "red", "TRUE" = "steelblue"),
                      name = "Prediction",
                      labels = c("Incorrect", "Correct")) +
    labs(title = "Confidence distribution by prediction correctness",
         x = "Confidence Score", y = "Frequency") +
    theme_classic()
  
  grid.arrange(p6, p7, ncol = 2)
  
  # Confidence statistics
  conf_stats <- cls_detailed %>%
    group_by(correct) %>%
    summarise(
      mean_confidence = mean(confidence, na.rm = TRUE),
      median_confidence = median(confidence, na.rm = TRUE),
      min_confidence = min(confidence, na.rm = TRUE),
      max_confidence = max(confidence, na.rm = TRUE),
      .groups = 'drop'
    ) %>%
    mutate(correct = ifelse(correct, "Correct", "Incorrect"))
  
  kable(conf_stats,
        caption = "Confidence score statistics by prediction correctness",
        digits = 3,
        align = "l")
}

```


## Conclusions and Recommendations

### Key Findings

1. **Detection Performance**: The YOLOv11n detection model achieved strong performance in dipteran detection with a mAP50-95 value of 0.99 on the test set.

2. **Classification Performance**: The YOLOv11n-cls classification model achieved an overall accuracy of 1 in genus identification.

3. **Training Efficiency**: Both models converged effectively within their respective training epochs, with both modela requiring 30 epochs.

## Technical specifications

```{r tech_specs}
# System information
tech_specs <- data.frame(
  Component = c(
    "YOLO Version",
    "Framework", 
    "Python Environment",
    "Image Input Size",
    "Inference Device",
    "Model Weights Format"
  ),
  Specification = c(
    "YOLOv11 (Ultralytics)",
    "PyTorch",
    "Python 3.8 with Ultralytics",
    "640x640 pixels",
    "GPU/CPU Compatible", 
    "PyTorch (.pt)"
  )
)

kable(tech_specs,
      caption = "Technical specifications",
      align = "l")
```

---

*Report generated using `r R.version.string`*

```{r session_info, include=FALSE}
# Capture session information for reproducibility
sessionInfo()
```